---
title: "Marathon_Visualisations"
author: "Brecht Dewilde"
date: "30-3-2020"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(tidyverse)
library(cowplot)
library(ggpubr)
library(gridExtra)
source("~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/function_library.R")
theme_set(theme_classic())
pic.path <- "~/Github/Predicting-the-olympic-games/marathon/IDA/Pictures/"
```

```{r load data, include=FALSE}
load("~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/Data/Rda/marathonM_extended.RData")
load("~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/Data/Rda/marathonF_extended.RData")
```

### Analysis of the Olympic marathon participants
We only use data from 1960 onwards <- Why?
```{r echo=FALSE, message=FALSE, warning=FALSE}
file = "~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/Data/athlete_events.csv"
fdata <- read.csv(file, stringsAsFactors = F)
marathon <-filter(fdata, grepl("Marathon", fdata$Event)) %>% filter(Year >= 1960)
men <- marathon %>% filter(Sex == "M") %>% count(Year)
men$Sex <- rep("M", nrow(men))
female <- marathon  %>% filter(Sex == "F") %>% count(Year)
female$Sex <- rep("F", nrow(female))
temp <- rbind(men,female)

p1 <- ggplot(temp, aes(x = Year, y =  n, shape = Sex, color = Sex)) + geom_point() +  scale_color_manual(values = c("magenta3", "black")) + 
  labs(title="Evolution of the number of athletes") + geom_smooth(span = 0.3, se = F, method = "loess")  + 
  scale_x_continuous(name = "Year", breaks = seq(1960, 2016,4)) + 
  scale_y_continuous(name = "Race participants", breaks = seq(40,160,10))
backup <- temp
rm("fdata", "female", "men", "temp", "file")
```

Number of athletes distributed over the sub-regions
```{r echo=FALSE, message=FALSE, warning=FALSE}
file = "~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/Data/noc_regions.csv"
noc_country <- read.csv(file)[,c(1,2)]
marathon <- merge(x = marathon, y = noc_country, all.x = TRUE)
marathon <- marathon[,c("Name", "Sex", "Age", "Height", "Weight", "Year", "City",  "Medal", "NOC", "region")]
colnames(marathon)[10] <- "Nationality"

file = "~/Github/Predicting-the-olympic-games/marathon/Data c&c/R/Data/countries_regions.csv"
countries.regions <- read.csv(file, stringsAsFactors = FALSE)[ ,c(1, 3, 7)]
colnames(countries.regions) <- c("Nationality", "NOC", "Sub-Region")
temp <- merge(marathon, countries.regions, by = "Nationality", all.x = TRUE)
temp <- merge(temp, countries.regions, by.x = "NOC.x", by.y = "NOC", all.x = TRUE)
temp$SubRegion <- ifelse(is.na(temp$`Sub-Region.x`), temp$`Sub-Region.y`, temp$`Sub-Region.x`)
subRegions <- temp[,c("Sex", "Age", "Height", "Weight", "Year", "SubRegion")]
subRegions$AggSubRegion <- unlist(lapply(subRegions$SubRegion, regionAggregator))
men <- subRegions %>% filter(Sex == "M") %>% group_by(AggSubRegion) %>% count(Year)
men <- na.omit(men)
female <- subRegions  %>% filter(Sex == "F") %>% group_by(AggSubRegion) %>% count(Year)
female <- na.omit(female)
rm("noc_country", "temp", "file")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
p2 <- ggplot(men, aes(x = Year, y =  n, color = AggSubRegion)) + geom_smooth(se = F) + 
 labs(title="Country participation", subtitle = "Men") +
  scale_x_continuous(name = "Year", breaks = seq(1960, 2016,8)) + 
  scale_y_continuous(name = "Race participants") + theme(legend.position="bottom", legend.title = element_blank()) + 
   scale_color_manual(values = c("darkgreen", "deeppink2", "darkblue", "brown4", "burlywood4", "antiquewhite2"))

p3 <- ggplot(female, aes(x = Year, y =  n, color = AggSubRegion)) + geom_smooth(se = F) + 
 labs(title="Country participation", subtitle = "Women") +
  scale_x_continuous(name = "Year", breaks = seq(1960, 2016,4)) + 
  scale_y_continuous(name = "Race participants") + theme(legend.position="bottom", legend.title = element_blank()) +
  scale_color_manual(values = c("darkgreen", "deeppink2", "darkblue", "brown4", "burlywood4", "antiquewhite2"))
```

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
save.path <- paste0(pic.path,"Participants.png")
ggarrange(p1, ggarrange(p2, p3, ncol = 2, common.legend = T,legend = "bottom"), nrow = 2) 
ggsave(save.path, dpi = "retina", height = 8, width = 12)
rm("p1", "p2", "p3")
```

### Analyis of the physical characteristics of the athletes
```{r echo=FALSE, message=FALSE, warning=FALSE}
output.men <- cor.test(marathonM$bmi, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 4), ")")
output.female <- cor.test(marathonF$bmi, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 5), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

test <- rbind(marathonM, marathonF)
bmi <- ggplot(test, aes(bmi, TimeMins, color = Sex)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y~x, se = F)  + 
  labs(title="BMI", subtitle = full )  + 
   scale_color_manual(values = c("magenta3", "black"))
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE}
output.men <- cor.test(marathonM$Height, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 4), ")")
output.female <- cor.test(marathonF$Height, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 5), ")")
full <- paste("Men:", cM, "\nWomen:", cF)
test <- rbind(marathonM, marathonF)
height <- ggplot(test, aes(Height, TimeMins, color = Sex)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y~x, se = F)  + 
  labs(title="Height", subtitle = full)  + 
   scale_color_manual(values = c("magenta3", "black")) + xlab("Height (cm)")
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE}
output.men <- cor.test(marathonM$Weight, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 4), ")")
output.female <- cor.test(marathonF$Weight, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 5), ")")
full <- paste("Men:", cM, "\nWomen:", cF)
test <- rbind(marathonM, marathonF)
weight <- ggplot(test, aes(Weight, TimeMins, color = Sex)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y~x, se = F)  + 
  labs(title="Weight", subtitle = full )  + 
   scale_color_manual(values = c("magenta3", "black")) + xlab("Weight (kg)")
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE}
output.men <- cor.test(marathonM$Age, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 2), ")")
output.female <- cor.test(marathonF$Age, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 2), ")")
full <- paste("Men:", cM, "\nWomen:", cF)
test <- rbind(marathonM, marathonF)
age <- ggplot(test, aes(Age, TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="Age", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black"))
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE}
statistics = c("Mean", "Std")
BMI = c(round(mean(marathonM$bmi), 2), round(sd(marathonM$bmi),2)) 
Height = round(c(mean(marathonM$Height), sd(marathonM$Height)),2)
Weight = round(c(mean(marathonM$Weight), sd(marathonM$Weight)),2)
Age = round(c(mean(marathonM$Age), sd(marathonM$Age)),2)
SummaryTableM <- data.frame(statistics, BMI, Height, Weight, Age)
colnames(SummaryTableM)[1] <- c("Men Statistics")

BMI = round(c(mean(marathonF$bmi), sd(marathonF$bmi)),2)
Height = round(c(mean(marathonF$Height), sd(marathonF$Height)),2)
Weight = round(c(mean(marathonF$Weight), sd(marathonF$Weight)),2)
Age = round(c(mean(marathonF$Age), sd(marathonF$Age)),2)
SummaryTableF <- data.frame(statistics, BMI, Height, Weight, Age)
colnames(SummaryTableF)[1] <- c("Women Statistics")

table.m <- ggtexttable(SummaryTableM, rows = NULL, 
                        theme = ttheme(base_size = 7.5))
table.f <- ggtexttable(SummaryTableF, rows = NULL,
                        theme = ttheme(base_size = 7.5))
```


```{r fig.height=10, fig.width=8}
figure <- ggarrange(bmi, height, weight, age, table.m, table.f, nrow = 3, ncol = 2, common.legend = T,legend = "top") 

save.path <- paste0(pic.path,"Physical characteristics.png")
annotate_figure(figure,
               top = text_grob("Physical characteristics", color = "black", face = "bold", size = 14))
ggsave(save.path, dpi = "retina", height = 8, width = 12)

```


### Analysis of target variable 'Race Time'
1) average time trend <- difference between male and female
```{r echo=FALSE, message=FALSE, warning=FALSE}
avgMen <- marathonM %>% group_by(Year) %>% summarise(avgTime = mean(TimeMins))
avgMen$Sex <- rep("M", nrow(avgMen))
avgFem <- marathonF %>% group_by(Year) %>% summarise(avgTime = mean(TimeMins))
avgFem$Sex <- rep("F", nrow(avgFem))
temp <- rbind(avgMen, avgFem)

pic <- ggplot(temp, aes(x = Year, y = avgTime, shape = Sex, color = Sex)) +
scale_y_continuous(name = "Finish time", 
                   breaks = seq(130,170, 10), 
                   labels = c("2:10:00", "2:20:00", "2:30:00", "2:40:00", "2:50:00"), 
                   limits = c(130, 170)) + 
  geom_hline(aes(yintercept = 130),  linetype = "dashed" , colour= 'blue') +    
  geom_hline(aes(yintercept = 140),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 150),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 160),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 170),  linetype = "dashed" , colour= 'blue') +  
  scale_x_continuous(name = "Year", breaks = seq(1960, 2016,4)) +
  geom_smooth(se = F, span = 0.5) +  scale_color_manual(values = c("magenta3", "black")) + labs(title="Finish time trend", subtitle = "Average finish time for each year")

save.path <- paste0(pic.path,"FinishTimeTrend.png")
pic
ggsave(save.path, dpi = "retina", height = 8, width = 12)


rm("fdata", "female", "men", "temp", "file", "avgFem", "avgMen", "fem")
```
2) fastest time <- difference between male and female
```{r echo=FALSE, message=FALSE, warning=FALSE}
minMen <- marathonM %>% group_by(Year) %>% summarise(minTime = min(TimeMins))
minMen$Sex <- rep("M", nrow(minMen))
minFem <- marathonF %>% group_by(Year) %>% summarise(minTime = min(TimeMins))
minFem$Sex <- rep("F", nrow(minFem))
temp <- rbind(minMen, minFem)

fastestsMen <- marathonM[match(min(marathonM$TimeMins), marathonM$TimeMins), c("Name", "Time", "TimeMins", "Year")]
fastestsFem <- marathonF[match(min(marathonF$TimeMins), marathonF$TimeMins), c("Name", "Time", "TimeMins", "Year")]



pic <- ggplot(temp, aes(x = Year, y = minTime, shape = Sex, color = Sex)) +  geom_line() +  #geom_smooth(se = F, span = 0.5) + 
scale_y_continuous(name = "Finish time", 
                   breaks = seq(120,160, 10), 
                   labels = c("2:00:00", "2:10:00", "2:20:00", "2:30:00", "2:40:00"), 
                   limits = c(120, 160)) + 
  geom_point(aes(x = fastestsMen$Year, y = fastestsMen$TimeMins), colour = "black") + 
  geom_point(aes(x = fastestsFem$Year, y = fastestsFem$TimeMins)) +
  geom_hline(aes(yintercept = 130),  linetype = "dashed" , colour= 'blue') +    
  geom_hline(aes(yintercept = 140),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 150),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 160),  linetype = "dashed" , colour= 'blue') +  
  geom_hline(aes(yintercept = 170),  linetype = "dashed" , colour= 'blue') +  
  scale_x_continuous(name = "Year", breaks = seq(1960, 2016,4)) +
  scale_color_manual(values = c("magenta3", "black")) + labs(title="Winner time trend", subtitle = "Winners finish time for each year") +
  annotate("segment", 
           y = c(fastestsMen$TimeMins - 2, fastestsFem$TimeMins + 7.5), x = c(2005, 2013), 
           yend =  c(fastestsMen$TimeMins - 0.3 , fastestsFem$TimeMins + 0.4), xend = c(fastestsMen$Year, fastestsFem$Year),
          arrow = arrow(angle = 20, length = unit(2, "mm"), type = "closed")) + 
          annotate(geom="text", x = 1999, y = fastestsMen$TimeMins  - 2, 
                   label= paste("Samuel Wanjiru\n", fastestsMen$Time), color="red") + 
           annotate(geom="text", x = 2013, y = fastestsFem$TimeMins + 10, 
                  label= paste("Tiki Gelena\n", fastestsFem$Time), color="red")

save.path <- paste0(pic.path,"WinnerTimeTrend.png")
pic
ggsave(save.path, dpi = "retina", height = 8, width = 12)
```

### Analyis of the demographic influences
```{r}
# plot: influence between gdp and race time
output.men <- cor.test(marathonM$GDP, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 11), ")")
output.female <- cor.test(marathonF$GDP, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 8), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

marathon <- rbind(marathonM, marathonF)
gdp <- ggplot(marathon, aes(log(GDP), TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="GDP", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black"))
```

```{r}
output.men <- cor.test(marathonM$Population, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 11), ")")
output.female <- cor.test(marathonF$Population, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 9), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

marathon <- rbind(marathonM, marathonF)
pop <- ggplot(marathon, aes(log(Population), TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="Population", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black"))

```


```{r fig.height=7, fig.width=6, message=FALSE, warning=FALSE}
wm <- map_data("world")
cc <- raster::ccodes()
# cc$MYCONTINENTS <- 

dat <- read.csv(text="state, RunTime
Other Europe, 158.36
Other Asian,   161.52
Other African,   155.96
Other North American,  152.15
Other Latin American,   163.45
Other Oceania,   155.31")


mappings <- c("Asia"="Other Asian", "Africa"="Other African", "North America" = "Other North American", 
              "Europe" = "Other Europe", "South America" = "Other Latin American", "Oceania" = "Other Oceania")

cc$MYCONTINENTS <- mappings[cc$continent]
cc <- left_join(cc, dat, by = c("MYCONTINENTS"="state"))

mappings <- c("UK"="United Kingdom", "USA"="United States") # You add the others here
cc$NAME[match(mappings, cc$NAME)] <- names(mappings)

wm <- left_join(wm, cc[,c("NAME","MYCONTINENTS", "RunTime")], by=c("region"="NAME"))
world.women <- ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=RunTime), wm, colour = NA) +
  coord_quickmap() + labs(subtitle = "Women") + xlab("") + ylab("") + labs(fill = "Continent Average Race Time (min)") + 
  theme(legend.direction = "horizontal", legend.position = c(0.5,-0.3),
  legend.key.size = unit(1.0, "cm"),
  legend.key.width = unit(1.5,"cm")) + 
    guides(fill = guide_colourbar(title.position = "top",
                                title.hjust = .5,
                                label.position = "bottom")) + 
  scale_fill_gradientn(colours = c("green", "orange", "red"),
                       values = c(0,0.5, 1))
```

```{r fig.height=7, fig.width=6}
wm <- map_data("world")
cc <- raster::ccodes()
# cc$MYCONTINENTS <- 


dat <- read.csv(text="state, RunTime
Other Europe, 141.14
Other Asian,   142.74
Other African,   139.27
Other North American,  138.28
Other Latin American,   144.44
Other Oceania,   141.68")


mappings <- c("Asia"="Other Asian", "Africa"="Other African", "North America" = "Other North American", 
              "Europe" = "Other Europe", "South America" = "Other Latin American", "Oceania" = "Other Oceania")

cc$MYCONTINENTS <- mappings[cc$continent]
cc <- left_join(cc, dat, by = c("MYCONTINENTS"="state"))

mappings <- c("UK"="United Kingdom", "USA"="United States") # You add the others here
cc$NAME[match(mappings, cc$NAME)] <- names(mappings)

wm <- left_join(wm, cc[,c("NAME","MYCONTINENTS", "RunTime")], by=c("region"="NAME"))
world.men <- ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=RunTime), wm, colour = NA) +
  coord_quickmap() + labs(subtitle = "Men")  + xlab("") + ylab("") + labs(fill = "Continent Average Race Time (min)") + 
  theme(legend.direction = "horizontal", legend.position = c(0.5,-0.3),
  legend.key.size = unit(1.0, "cm"),
  legend.key.width = unit(1.5,"cm")) + 
    guides(fill = guide_colourbar(title.position = "top",
                                title.hjust = .5,
                                label.position = "bottom")) + 
  scale_fill_gradientn(colours = c("green", "orange", "red"),
                       values = c(0,0.5, 1))
```
```{r}
#world.men, world.women, gpd, population
figure <- ggarrange(gdp, pop, world.men, world.women, nrow = 2, ncol = 2) 
fig <-annotate_figure(figure,
               top = text_grob("Demographic influences", color = "black", face = "bold", size = 14))
save.path <- paste0(pic.path,"Demographics.png")
fig
ggsave(save.path, dpi = "retina", height = 9.5, width = 12)
```

### Analysis of the Performance features
```{r}
# plot: influence between TimeScore and race time
output.men <- cor.test(marathonM$avgTimeScore, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 44), ")")
output.female <- cor.test(marathonF$avgTimeScore, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 38), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

marathon <- rbind(marathonM, marathonF)
timeScore<- ggplot(marathon, aes(avgTimeScore, TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="Time score", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black")) + xlab("Time score")
timeScore
```

```{r}
# plot: influence between TimeScore and race time
output.men <- cor.test(marathonM$avgPositionScore, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 10), ")")
output.female <- cor.test(marathonF$avgPositionScore, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 3), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

marathon <- rbind(marathonM, marathonF)
positionScore<- ggplot(marathon, aes(avgPositionScore, TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="Position score", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black")) + xlab("Position score")
positionScore
```


```{r}
# plot: influence between TimeScore and race time
output.men <- cor.test(marathonM$avgPerformanceScore, marathonM$TimeMins, method =  "spearman", exact = F)
cM <- paste0(round(output.men$estimate, 2), " (p:", round(output.men$p.value, 38), ")")
output.female <- cor.test(marathonF$avgPerformanceScore, marathonF$TimeMins, method =  "spearman", exact = F)
cF <- paste0(round(output.female$estimate, 2), " (p:", round(output.female$p.value, 34), ")")
full <- paste("Men:", cM, "\nWomen:", cF)

marathon <- rbind(marathonM, marathonF)
performanceScore<- ggplot(marathon, aes(avgPerformanceScore, TimeMins, color = Sex)) + 
        geom_point() + 
        geom_smooth(method = "lm", formula = y~x, se = F)  + 
        labs(title="Performance score", subtitle = full)  + 
        scale_color_manual(values = c("magenta3", "black")) + 
        annotate("text", x = 1200, y = 210, label = "Performance score = Time score + Position score", size = 4, color = "brown4") + xlab("Performance score") +  theme(legend.position = "none")
performanceScore
```

```{r}
save.path <- paste0(pic.path,"Performance.png")
ggarrange(performanceScore, ggarrange(timeScore, positionScore, ncol = 2, common.legend = T,legend = "bottom"), nrow = 2) 
ggsave(save.path, dpi = "retina", height = 8, width = 12)
```

